``Interestingness'' in Mathematical Theory Exploration

### 2019-04-29 PhDThesis/litreview_premiseselection.tex
Clean up

One fruitful application of machine learning methods to formal mathematics is
\emph{relevance filtering}, which is used for \emph{premise selection} in
automated theorem provers~\cite{kuhlwein2012overview}.

Relevance filtering simplifies a proof search problem by removing from
consideration those clauses (axioms, definitions, lemmas, etc.) which are deemed
\emph{irrelevant}. The technique is used in Sledgehammer during its translation
of Isabelle/HOL theories to statements in first order logic: rather than
translating the entire theory, only a sub-set of relevant clauses are
included. This reduces the size of the problem and speeds up the proof search,
but it creates the new problem of determining when a clause is relevant: how do
we know what will be required, before we have the proof?

Sledgehammer's initial approach, known as \textsc{MePO} (from
\emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based
on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is
the number of symbols in the clause and $m$ is the number which are relevant).
Initially, the relevant symbols are those which occur in the goal, but whenever
a clause is found which scores more than a particular threshold, all of its
symbols are then also considered relevant. There are other heuristics applied
too, such as increasing the score of user-provided facts (e.g. given by keywords
like \texttt{using}), locally-scoped facts, first-order facts and
rarely-occuring facts. To choose $r$ relevant clauses for an ATP invocation, we
simply order the clauses by decreasing score and take the first $r$ of them.

Recently, a variety of alternative algorithms have also been
investigated~\cite{kuhlwein2013mash,hoder2011sine,urban2013blistr},
%                  MaSH             SInE,         BliStr
dubbed , \textsc{SInE}, \textsc{BliStr}, \textsc{HOLyHammer},
\textsc{MoMM}, \textsc{SNoW}, \textsc{MPTP 0.2}, \textsc{MaLARea} and
\textsc{MaLARea SG1}. Each one combines an ``off the shelf'' machine learning
algorithm with a custom feature extraction approach, is trained using a corpus
of existing theorems and their proofs, and evaluated by trying to re-prove a
selection of unseen theorems from the corpus (i.e. given only the theorem
statement and possible lemmas, not the existing proof).

For example the Machine Learning for SledgeHammer (\textsc{MaSH}) system
defines a notion of ``visibility''~\cite{kuhlwein2013mash}. This is essentially
a dependency graph of which theorems were used in the proofs of which other
theorems (theorems are represented as abstract sets of features, rather than
e.g. sentences or syntax trees). To select relevant clauses for a goal, the
set of clauses which are ``visible'' from the goal's components is generated;
this is further reduced by (an efficient approximation of) a naive Bayes
algorithm.

Premise selection has also been applied to the re-proving problem in other
systems, such as the Multi-Output Ranking (\textsc{MOR}) tool, which uses a
support vector machine (SVM) to select relevant axioms from the Mizar
Mathematical Library for use by the Vampire ATP system~\cite{alama2014premise}.
%TODO: describe the kernel, as that's the interesting bit.

The premise selection problem for theorem proving is related to the signature
selection problem for theory exploration, in that both are trying to pick a
small number of terms (depending on the speed of the subsequent algorithm and
how long the user is willing to wait) from a large set, to optimise the outcome
of that algorithm (e.g. finding a proof quickly; or finding many interesting
relationships quickly). The major difference is that premise selection has a
particular ``target'' theorem statement, against which to compare each
term. This is not the case for theory exploration, so we must consider the
interactions of each term with all of the others when deciding what to select.

>>>>

\emph{Relevance filtering} takes a specified ``target'' (such as the current
goal of a theorem prover), a large list of inputs (e.g. existing lemmas from a
theory) and produces a sub-set of those inputs which are predicted to be useful
or relevant to the target. The example of choosing a sub-set of lemmas to help
prove a theorem is known as \emph{premise
  selection}~\cite{kuhlwein2012overview}.

Relevance filtering simplifies such proof search by focusing on only those
clauses (axioms, definitions, lemmas, etc.) which are deemed relevant, shrinking
the search space. The technique is used in \textsc{Sledgehammer} during its
translation of Isabelle/HOL theories to statements in first order logic: rather
than translating the entire theory, only a sub-set of relevant clauses are
included. This speeds up the proof search, but it creates the new problem of
determining when a clause is relevant: how do we know what will be required,
before we have the proof?

The initial approach used in \textsc{Sledgehammer}, known as \textsc{MePO} (from
\emph{Meng-Paulson} \cite{meng2009lightweight}), gives each clause a score based
on the proportion $m / n$ of its symbols which are ``relevant'' (where $n$ is
the number of symbols in the clause and $m$ is the number which are relevant).
Initially, the relevant symbols are those which occur in the goal, but whenever
a clause is found which scores more than a particular threshold, all of its
symbols are then also considered relevant. There are other heuristics applied
too, such as increasing the score of user-provided facts (given by keywords
like \texttt{using}), locally-scoped facts, first-order facts and
rarely-occuring facts. To choose $r$ relevant clauses for an ATP invocation, we
simply order the clauses by decreasing score and take the first $r$ of them.

Recently, a variety of alternative algorithms have also been
investigated~\cite{kuhlwein2013mash,hoder2011sine,urban2013blistr,gauthier2015premise,alama2014premise,carlson1999snow,kaliszyk2014machinelearner,kaliszyk2015femalecop}.
%                  MaSH             SInE,         BliStr          HOLyHammer          MOR,             SNoW,           MaLARea
Each combines an ``off the shelf'' machine learning algorithm (e.g. a
naive-Bayes classifier, support vector machines, etc.) with a custom feature
extraction approach; is trained using a corpus of existing theorems and their
proofs; and evaluated by trying to \emph{re-prove} a selection of unseen
theorems from the corpus (i.e. given only the theorem statement and possible
lemmas, not the existing proof).

For example the ``Machine Learning for SledgeHammer'' (\textsc{MaSH}) system
defines a notion of ``visibility''~\cite{kuhlwein2013mash} for determining how
relevant one statement is to another. Visibility is essentially a dependency
graph of which theorems were used in the proofs of which other theorems
(represented as abstract sets of features, rather than concrete sentences or
syntax trees). To select relevant clauses for a goal, the set of clauses which
are ``visible'' from the goal's components is generated; and this is filtered
further by (an efficient approximation of) a naive-Bayes algorithm.

Premise selection has also been applied to the re-proving problem in other
systems, such as the Multi-Output Ranking (\textsc{MOR}) tool, which uses a
support vector machine (SVM) to select relevant axioms from the Mizar
Mathematical Library (MML) for use by the Vampire ATP
system~\cite{alama2014premise}. SVM requires a ``kernel function'', effectively
a similarity measure between formulas. The \textsc{MOR} tool measures similarity
based on sub-terms which two formulas have in common: first a list of all $m$
terms appearing in the input space is extracted, and each formula is converted
to an $m$-dimensional binary vector, with the $n$th element of the vector
indicating whether the $n$th term of the input space occurs in this formula. The
kernel used by \textsc{MOR} is a Gaussian function of the dot-products of
these binary vectors (parameterised by a hyperparameter $\sigma$, chosen via
grid search).

The premise selection problem for theorem proving is related to the signature
selection problem for theory exploration, in that both are trying to pick a
small number of terms (depending on the speed of the subsequent algorithm and
how long the user is willing to wait) from a large set, to optimise the outcome
of that algorithm (e.g. finding a proof quickly; or finding many interesting
relationships quickly). The major difference is that premise selection has a
particular ``target'' theorem statement, against which to compare each
term. Indeed, machine learning approaches like \textsc{MOR} learn a separate
ranking function for each target formula.

This is not the case for theory exploration, which must consider the
interactions of each term with all of the others when deciding what to select.

### 2019-04-29 PhDThesis/litreview_testing.tex
Clean up

Automated testing is a common method of software quality assurance. Typical
industry standards (outside of specialised domains like those which are
life-critical) include \emph{unit testing} \iffalse \cite{TODO} \fi: asserting
that components of a system behave as expected on a fixed set of inputs. In
Haskell, some of the most popular testing frameworks go beyond this to instead
perform \emph{property checking}. A ``property'', in this sense, is a statement
about a program's components, which may include free variables. We can think of
a unit test as proving an \emph{existential} property: that the given statement
holds for at least one assignment of the variables, and those values are stated
as part of the unit test. However, the phrase ``property checking'' is typically
reserved for \emph{universal} properties: those which should hold \emph{for all}
values of the variables.

Universal properties are typically encoded in Haskell as \emph{functions},
accepting values for the free variables as arguments, and returning a boolean
value (perhaps adorned with useful metadata). If the type of the variables is
finite (for example booleans), it might be feasible to check all possible
values. Typically this is not feasible, especially when multiple variables are
involved, or the input types contain an infinite number of values (for example
lists). Property checkers avoid this by only testing a sub-set of values, in the
hope that a counterexample will be found (if one exists). They work by:

\begin{enumerate}
\item Generating suitable values for the function's arguments
\item Calling the function with these values
\item Inspecting the return value
\item Repeating until some stop condition is satisfied
\end{enumerate}.

@@

\textsc{QuickCheck}~\cite{claessen2011quickcheck} popularised the use of
property checking in Haskell, and has subsequently been ported to many other
programming languages. \textsc{QuickCheck} generates input data \emph{randomly},
either using explicit generators (via a function called \texttt{forAll}), or by
choosing a default generator for types implementing its \texttt{Arbitrary}
typeclass (the recent \texttt{hedgehog} package works in a similar way\iffalse
\cite{TODO} \fi).

Random testing is effective at thoroughly exercising the branches of a program,
since inputs may be quite large and cover many cases, especially for functions
which recurse on each \emph{sub-}expressions of their input as well. The output
of a random generator will also vary across runs, rather than checking the same
values every time. Hence every invocation of a property checker, e.g. as part of
a development or debugging workflow, allows more of the input space to be
tested. However, there are two common problems with random property checking.

Firstly we must be able to generate random values, which can be tricky for
recursive datatypes. Consider a type of binary trees, with constructors
\texttt{Leaf~::~Tree} and \texttt{Node~::~Tree~->~Tree~->~Tree}. We might
generate a \texttt{Tree} by picking uniformly between these two constructors
and, if a \texttt{Node} was picked, recursing to generate the two
sub\texttt{Tree}s. Yet this simple scheme will generate \texttt{Tree}s whose
expected size is \emph{unbounded}! This is due to the 50\% chance of generating
a \texttt{Node}, which requires two recursive calls; hence we can expect each
call to recurse $0.5~\times~2~=~1$ more time on average, and that call will also
recurse once more on average, and so on; this can easily exhaust all available
memory. Many automated data generators, such as those from Haskell's
\texttt{derive} package, suffer from this problem. It can be avoided by biasing
the generators (in the \texttt{Tree} example we would produce a \texttt{Leaf}
more than half of the time), but this bias compounds at each level of recursion,
limiting testing to very small values. More sophisticated solutions exist, such
as the \texttt{boltzmann-sampling} package (see \cite{duchon2004boltzmann}) and
even domain-specific languages~\cite{lampropoulos2017beginner}, but all of these
add complications to the property checking process.

The other problem with random generators is that it can be difficult to
understand \emph{why} a randomly generated value violated some property.
\textsc{QuickCheck} tackles this in two ways: generators have access to a
``size'' parameter, with the intention that this be used to decide how large the
output values should be (e.g. the expected length of a list, or number of nodes
in a tree). The value of the size parameter begins low and increases as more
tests pass, such that failures are likely to be triggered by small, more easily
understood counterexamples first. Additionally, a mechanism for ``shrinking'' is
provided: users can supply a \texttt{shrink} function along with their
generator, which is given any counterexample found during testing and returns a
list of smaller values. In the above example we might shrink a \texttt{Node} by
returning each sub-tree, and return an empty list when asked to shrink a
\texttt{Leaf}. If one of these smaller values is also a counterexample, it too
is shrunk and the process repeats, otherwise it is discarded and the next shrunk
value is tried. When a value cannot be shrunk any more, or if none of its
shrunken versions are counterexamples, that value is presented to the user in
the hope that it is simple enough to understand.

The \textsc{SmallCheck} framework takes a different approach: \emph{enumerating}
values (either using a user-provided enumeration function or using its
\texttt{Serial} typeclass), rather than generating values
randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
the \texttt{testing-feat} package \iffalse \cite{}TODO} \fi). In the
\texttt{Tree} example this might check \texttt{Leaf}, then \texttt{Node Leaf
  Leaf}, then \texttt{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
order of ``size'' ensures that counterexamples are always small, that tests are
not repeated with the same values and that memory is not exhausted (since the
stopping condition will trigger long before the enumeration reaches such large
values). The disadvantage is that large inputs (which exercise properties,
especially recursive ones, the most) may occur so late in the enumeration that
testing never reaches them.

The \textsc{LazySmallCheck} framework is similar in spirit to
\textsc{SmallCheck}, but utilises Haskell's lazy evaluation to avoid generating
values up front~\cite{reich2013advances}. Instead, a partial value is generated
which will trigger an exception if evaluated. Since catching a Haskell exception
is an \texttt{IO} action, pure properties which inspect such values will always
be aborted, handing control back to \textsc{LazySmallCheck} which checks the
same property on refined versions of the previous value. For example, a lazily
generated \texttt{Tree} may initially be equivalent to $\bot$. If the property
aborts, it will be run again with each possible refinement of $\bot$: in this
case inputs equivalent to \texttt{Leaf} and \texttt{Node~$\bot_1$~$\bot_2$}. If
the latter aborts due to inspecting $\bot_1$, then $\bot_1$ will be refined to
give inputs \texttt{Node~Leaf~$\bot_2$} and
\texttt{Node~(Node~$\bot_3$~$\bot_4$)~$\bot_2$} (and likewise if $\bot_2$ had
caused the exception). This refinement continues until the property no longer
aborts or until a predefined depth limit is reached. If the property returns a
true value when given such partially-defined input then it must succeed for
\emph{all} refinements of that input (since the output was determined without
evaluating the undefined sub-terms). If the property returns a false value, the
input is a \emph{minimal} counterexample, since it only specifies those
sub-terms which are required to violate the property; any undefined sub-terms
are irrelevant to the failure.

>>>>

Some prominent tools for mathematical theory exploration, like \qspec{}, were
originally developed in the context of \emph{property checking} (although for
\emph{discovering} software properties, rather than checking them), and only
later applied more directly to fields like formal mathematics (e.g. via the
\hipster{} tool for Isabelle/HOL). In particular, the logical domain used by
\qspec{} (inductive logic with higher-order functions) comes directly from its
application to Haskell; and its search algorithm makes heavy use of the
\qcheck{} property checker. The boundary between these two fields is hence
rather fuzzy, and cross-pollination of techniques is a compelling target for
research.

Property checking attempts to establish the truth or falsity of a statement (a
``property''), usually involving components of a software system and, crucially,
may contain free variables. \emph{Unit testing} is a common method of software
quality assurance (outside of specialised domains like those which are
life-critical), which can be characterised as proving \emph{existential}
properties: each test includes concrete values for any variables, and checking
is simply a matter of evaluating the code to get a boolean value (pass or fail).
The phrase ``property checking'' is typically reserved for \emph{universal}
properties: those which should hold \emph{for all} values of the variables.
This is a popular method of quality assurance for Haskell software, giving rise
to a number of packages which approach this challenge in different ways.

Universal properties are typically encoded in Haskell as \emph{functions},
accepting values for the free variables as arguments, and returning a boolean
value (perhaps adorned with useful metadata). If the type of the input is finite
(for example booleans), it might be feasible to check all possible
values. Typically this is not the case, especially when multiple variables are
involved, or the types have an infinite number of members (for example
lists). Property checkers avoid this by only testing a sub-set of values, in the
hope that a counterexample will be found (if one exists). They work by:

\begin{enumerate}
\item Generating suitable values for the function's arguments
\item Calling the function with these values
\item Inspecting the return value
\item Repeating until some stop condition is satisfied
\end{enumerate}

@@

\textsc{QuickCheck}~\cite{claessen2011quickcheck} popularised the use of
property checking in Haskell, and has subsequently been ported to many other
programming languages. \textsc{QuickCheck} generates input data \emph{randomly},
either using explicit generators (via a function called \hs{forAll}), or by
choosing a default generator for types implementing its \hs{Arbitrary}
typeclass (the recent \texttt{hedgehog} package works in a similar way\iffalse
\cite{TODO} \fi).

Random testing is effective at thoroughly exercising the branches of a program,
since inputs may be quite large and cover many cases, especially for functions
which recurse on each \emph{sub-}expression of their input as well. The output
of a random generator will also vary across runs, rather than checking the same
values every time. Hence every invocation of a property checker, e.g. as part of
a development or debugging workflow, allows more of the input space to be
tested. However, there are two common problems with random property checking.

Firstly we must be able to generate random values, which can be tricky for
recursive datatypes. Consider a type of binary trees, with constructors
\hs{Leaf :: Tree} and \hs{Node :: Tree -> Tree -> Tree}. We might
generate a \hs{Tree} by picking uniformly between these two constructors
and, if a \hs{Node} was picked, recursing to generate the two
sub\hs{Tree}s. Yet this simple scheme will generate \hs{Tree}s whose
expected size is \emph{unbounded}! This is due to the 50\% chance of generating
a \hs{Node}, which requires two recursive calls; hence we can expect each
call to recurse $0.5~\times~0~+~0.5~\times~2~=~1$ more time on average, and that call will also
recurse once more on average, and so on; this can easily exhaust all available
memory. Many automated data generators, such as those from Haskell's
\texttt{derive} package, suffer from this problem. It can be avoided by biasing
the generators (in the \hs{Tree} example we would produce a \hs{Leaf}
more than half of the time), but this bias compounds at each level of recursion,
limiting testing to exponentially small values. More sophisticated solutions
exist, such as the \texttt{boltzmann-sampling} package (see
\cite{duchon2004boltzmann}) and even domain-specific
languages~\cite{lampropoulos2017beginner}, but all of these add complications to
the property checking process.

The other problem with random generators is that it can be difficult to
understand \emph{why} a randomly generated value violated some property.
\textsc{QuickCheck} tackles this in two ways: generators have access to a
``size'' parameter, with the intention that this be used to decide how large the
output values should be (e.g. the expected length of a list, or number of nodes
in a tree). The value of the size parameter begins low and increases as more
tests pass, such that failures are likely to be triggered by small, more easily
understood counterexamples first. Additionally, a mechanism for ``shrinking'' is
provided: users can supply a \hs{shrink} function along with their generator,
which is given any counterexample found during testing and returns a list of
smaller values. In the above example we might shrink a \hs{Node} by returning
each sub-tree, and return an empty list when asked to shrink a \hs{Leaf}. If one
of these smaller values is also a counterexample, it too is shrunk and the
process repeats, otherwise it is discarded and the next element of the list is
tried. When a value cannot be shrunk any more, or if none of its shrunken
versions are counterexamples, that value is presented to the user in the hope
that it is simple enough to understand.

The \textsc{SmallCheck} framework takes a different approach: \emph{enumerating}
values (either using a user-provided enumeration function or using its
\hs{Serial} typeclass), rather than generating values
randomly~\cite{runciman2008smallcheck} (a similar approach is also followed by
the \texttt{testing-feat} package \iffalse \cite{}TODO} \fi). In the
\hs{Tree} example this might check \hs{Leaf}, then \hs{Node Leaf
  Leaf}, then \hs{Node (Node Leaf Leaf) Leaf}, and so on. Enumerating in
order of ``size'' ensures that counterexamples are always small, that tests are
not repeated with the same values and that memory is not exhausted (since the
stopping condition will trigger long before the enumeration reaches such large
values). The disadvantage is that large inputs (which exercise properties,
especially recursive ones, the most) may occur so late in the enumeration that
testing never reaches them.

The \textsc{LazySmallCheck} framework is similar in spirit to
\textsc{SmallCheck}, but utilises Haskell's lazy evaluation to avoid generating
values up front~\cite{reich2013advances}. Instead, a partial value is generated
which will trigger an exception if evaluated. Since catching a Haskell exception
is an \hs{IO} action, pure properties which inspect such values will always
be aborted, handing control back to \textsc{LazySmallCheck} which refines the
previous value and tries again. For example, a lazily generated \hs{Tree} may
initially be equivalent to $\bot$ (an undefined value). If the property
aborts, it will be run again with each possible refinement of $\bot$: in this
case inputs equivalent to \hs{Leaf} and \hs{Node}~$\bot_1$~$\bot_2$. If
the latter aborts due to inspecting $\bot_1$, then $\bot_1$ will be refined to
give inputs \hs{Node Leaf}~$\bot_2$ and
\hs{Node (Node}~$\bot_3$~$\bot_4$\hs{)}~$\bot_2$ (and likewise if $\bot_2$
had caused the exception). This refinement continues until the property no
longer aborts or until a predefined depth limit is reached. If the property
returns a true value when given such partially-defined input then it must
succeed for \emph{all} refinements of that input (since the output was
determined without evaluating the undefined sub-terms). If the property returns
a false value, the input is a \emph{minimal} counterexample, since it only
specifies those sub-terms which are required to violate the property; any
undefined sub-terms are irrelevant to the failure.
